---
title: "ggprovenance HowTo"
author: "Martin Rittner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggprovenance HowTo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1. Introduction {#intro}
Package `provenance` provides functions for the visualisation of typical data utilised in sediment provenance analysis in the geosciences, as well as helper functions aiding data analysis. Plotting takes advantage of the framework provided by the `ggplot2` package, and the output plots are ggplot objects, which allows for further modification with `ggplot2`'s functions and easy saving of the graphics. This document is meant for geoscientists who want to get step-by-step guidelines for using `provenance`'s plotting functions, and who might not be experienced users of the R language.  
The "workhorse functions" of the package are:

`plotKDE()` provides plotting of kernel density estimates (**KDE**s) of geochronological data for single or suites of samples

`plotMDS()` provides plotting of multidimensional scaling (**MDS**) maps for quick visual identification of trends and similar groups within a set of sample data

`plot3Way()` is designed to be very similar to `plotMDS()`, providing individual scaling (or **3-way MDS**) maps of data sets combining several different provenance proxies

Working with provenance data using the `ggprovenance` package in R typically involves three main steps:

1. load data files, filter and reformat input data for use
2. plot visualisation of data
3. save output

Of these steps, 2 usually is an iterative "trial-and-error" process to determine best graphical representation of the data, with a few additional calculations necessary here and there. Step 3 is, due to `ggplot2`'s functionality, a trivial call to `ggsave()`. The first item, loading the data, is usually the most involved and requires the user to write their own scripts for the purpose, as this can not easily be standardised over the wide range of possible data file formats and the individual requirements of each user. This step can be further subdivided into:

> 1a. loading data files  
> 1b. reformattiong data  
> 1c. adding information for visualisation

All basic steps will be presented in workable examples in [section 2](#workflow) of this document. [Section 3](#examples) illustrates the effects of different settings of the parameters given to the main plotting functions. [Section 4](#script) gives a basic generic script that should be easily adaptable to the user's needs. [Section 5](#tips) contains additional tips & tricks on plotting and data import.

\pagebreak

# Basic workflow {#workflow}
First, we need to load the package:
```{r message=FALSE}
library(ggprovenance)
```

## Loading data
For this document, we use example data files installed with `provenance`, but the principles would be the same for any other data file. Simple example data files are provided in the /inst subfolder of the package installation folder, these can serve as an example of how to prepare data. Alternatively, the data loading process can be adapted to the existing data files, and the data restructured in R, to obtain objects suitable for the plotting functions. The latter approach is preferrable, as it leaves the original data files untouched, and new and altered data is easily replotted by simply re-running the R script, instead of preparing intermediate data files by hand. See also section [Tips & Tricks](#import) for notes on how to import  lists of individual data files and MS Excel files.

### Loading data files
In the /inst folder, there is a simple example file of detrital zircon U-Pb ages. The individual measured ages are arranged column-wise, one column per sample, with the sample names in the first line.  
```{r, results='asis', echo=FALSE}
table<-read.xls(xls=system.file("extdata", "Tarim.xls", package="ggprovenance"),stringsAsFactors=FALSE)
knitr::kable(table[1:6,],format="markdown")
```

To load this data, you could use:
```{r}
agedata<-read.xls.flat(system.file("extdata", "Tarim.xls", package="ggprovenance"))
```

The data table is loaded with `read.csv()`, the result is a data frame. The default behaviour of R is to convert character vectors into factors (see R help); since `read.csv()` reads the (text) file with headers first, and more importantly in the case of reading Excel files (see [Tips & Tricks](#tips)) which are converted in csv files in an intermediate step in the background, numerical values can end up being interpreted as factors of strings that look like numbers, but cause nothing but headaches and confusion. For most cases discussed here, factors should be avoided.  
The functions within `provenance` generally work with lists, where each element of the list would be a vector of ages for one sample, respectively. Since data frames are a suclass of list, no further conversion is needed.  
Before plotting, we can have a look at some properties of the data:
```{r}
names(agedata)
```

We see there is data from three sampling sites, two of which are comprised of several subsamples, three potential source areas ("Area 1 - 3"), a river sample and a bedrock sample.
